---
title: "STAT6390_HW2_CongZhang"
author: "Cong Zhang"
date: "October 6, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = T)
```


1. Show that (algebraically) in the absence of censoring $\hat S_{{\tiny \mbox{KM}}}(t)=\hat S_{\mbox{e}}(t)$.

**Answer** 

$$\hat S_{{\tiny \mbox{KM}}}(t) = P(T>t) = P(T>t_{(0)}) \cdot  P(T>t_{(1)}|T>t_{(0)}) \cdot ... \cdot P(T>t|T>t_{(i)})$$
for a series of time intervals $0 = t_{0} < t_{1} < ... < t_{i} < t$ for some $i \le n$.

When there is no censoring, we have

$$\hat S_{{\tiny \mbox{KM}}}(t) = \prod_{t_{(i)} \le t} \frac{n_i - d_i}{n_i} = 1 \cdot \frac{n-1}{n} \cdot \frac{n-2}{n-1} \cdot ... \cdot \frac{ n_i -\sum_{j=1}^i d_j }{n_i} = \frac{ \sum_{i=1}^n I(T_i>t) }{n} = \hat S_{e}(t)$$



2. In the absence of censoring, show that the Greenwood Formula (page 30 on note 2) can be reduced to $$\frac{\hat S_{{\tiny \mbox{KM}}}(t) \times \{ 1 - \hat S_{{\tiny \mbox{KM}}}(t)\}}{n}.$$
You might assume there are no ties among the observations.

**Answer** 

$$\begin{aligned}
& \hat S_{{\tiny \mbox{KM}}}^2(t) \cdot \sum_{{t_{(i)} \le t}} \frac{d_i}{n_i \cdot (n_i - d_i)} \\
& = \hat S_{{\tiny \mbox{KM}}}^2(t) \cdot (\frac{1}{n_i - d_i} - \frac{1}{n_i} +\frac{1}{n_{i-1} - d_{i-1}} - \frac{1}{n_{i-1}} + ... + \frac{1}{n_1 - d_1} - \frac{1}{n_1}  ) \\ & = \hat S_{{\tiny \mbox{KM}}}^2(t) \cdot (\frac{1}{n_1 - \sum_{j=1}^i d_j} - \frac{1}{n_1 - \sum_{j=1}^{i-1} d_j} +\frac{1}{n_1 - \sum_{j=1-1}^i d_j} - \frac{1}{n_1 - \sum_{j=1}^{i-2} d_j} + ... + \frac{1}{n_1 - d_1} - \frac{1}{n_1}  ) \\ & =\hat S_{{\tiny \mbox{KM}}}^2(t) \cdot (\frac{1}{n_1 - \sum_{j=1}^i d_j}  - \frac{1}{n_1}  ) \\ & =\hat S_{{\tiny \mbox{KM}}}^2(t) \cdot \frac{\sum_{j=1}^i d_j}{n_1(n_1 - \sum_{j=1}^i d_j)} \\ & =\hat S_{{\tiny \mbox{KM}}}(t) \cdot \frac{n_1 - \sum_{j=1}^i d_j}{n_1} \cdot \frac{\sum_{j=1}^i d_j}{n_1} \cdot \frac{1}{n_1 - \sum_{j=1}^i d_j} \\ & =\hat S_{{\tiny \mbox{KM}}}(t) \cdot (1 - \frac{n_1 - \sum_{j=1}^i d_j}{n_1}) \cdot \frac{1}{n_1} \\ & =\frac{\hat S_{{\tiny \mbox{KM}}}(t) \times \{ 1 - \hat S_{{\tiny \mbox{KM}}}(t)\}}{n}
\end{aligned}$$


3. Consider the Leukemia data from the `survival` package: \label{aml}
    ```{r aml}
    library(survival)
    head(aml)
    ```
    In here, each row represent one patient. `aml` is the observed survival time, `status` is the censoring indicator (1 = event, 0 = censored), and `x` is the treatment indicator. We will ignore the treatment indicator for now.
    a. Plot the Kaplan-Meier survival curve for the data. \label{aml:surv}
    b. Add the Nelson-Aalen survival curve to the Kaplan-Meier plot from \eqref{aml:surv}.
    
    **Answer** 
    ```{r }
    library(survival)
    library(survMisc)
    library(knitr)
    km <- survfit(Surv(time, status) ~ 1, data = aml)
    cox <- coxph(Surv(time, status) ~ 1, data = aml)
    H0 <- basehaz(cox)
    plot(km, xlab = "Time", ylab = "Survival probability")
    lines(H0$time, exp(-H0$hazard), lty = 4, col = 2)
    legend("topright" , c("Kaplan-Meier", "Nelson-Aalon"), col = c(1, 2), lty = c(1, 4) )
    ```

4. The expected survival time for the Leukemia data in \#\eqref{aml} does not exist because the last observation is a censored event.
    An alternative is to lookInstead of looking at the expected survival time, an alternative is to look at the restricted mean survival time.
    Compute $E(T|T<161)$ based on the survival curve in \eqref{aml:surv}.
    
    **Answer** $E(T|T<161)$ = 36.36
    
    ```{r}
    survival:::survmean(km, rmean=161)
    ```
    
5. Let $N_i(t)$ be the number of events over time interval $(0, t]$ for the $i$th patient in \#\eqref{aml}. Let $N(t)=\sum_{i=1}^nN_i(t)$ be the aggregated counting process.
    a. Plot $N(t)$.
    b. Plot $M(t)$, where $M(t) = N(t) - \hat H(t)$ and $\hat H(t)$ is the Nelson-Aalen estimator for the cumulative hazard function. \label{Mt}
    
    * Note on \ref{Mt}: After giving some thought, I think it is more meaningful to plot $dM(t)=dN(t)- \hat h(t)dt$. Both plots will receive full credit for \ref{Mt}.
    
    **Answer**
    
    ```{r}
    library(ggplot2)
    library(dplyr)
    N <- Vectorize(function(t)
    {
        sum(aml[aml$time <= t, "status"])
    })
    
    HNA <- Vectorize(function(t)
    {
        if (H0$time[1] > t) 0 else H0[last(which(H0$time <= t)), "hazard"]
    })
    
    M <- function(t)
    {
        N(t) - HNA(t)
    }
    
    xlim <- c(0, max(aml$time))
    ylim <- c(0, 25)
    ggplot(data = data.frame(x = xlim), aes(x)) +
        stat_function(fun = N, aes(color = "N"), size = 1, n = 500) +
        stat_function(fun = M, aes(color = "M"), size = 1, n = 500, linetype = "dashed") +
        scale_x_continuous(name = "time",  limit = xlim) +
        scale_y_continuous(name = "value", limit = ylim, 
                           breaks =seq(ylim[1], ylim[2], length.out = 6)) +
        scale_color_manual(name = "functions",
                           values = c("N" = "skyblue", "M" = "mistyrose"),
                           breaks = c("N","M"),
                           labels = c("N(t)","M(t)")) +
        theme_bw()
    ```
    