---
title: Homework 2
author: Xinyu Zhang
output: pdf_document
---

\centering Due date: Thursday, October 11

```{R setup, include = FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

    ```{R load_library,echo=FALSE,message=FALSE}
    library(tidyverse)
    ```

1. Show that (algebraically) in the absence of censoring $\hat S_{{\tiny \mbox{KM}}}(t)=\hat S_{\mbox{e}}(t)$.

    $Assume\ m=max\{i:t_{(i)}\leq{t}\},\ then:\newline$

    $\hat S_{{\tiny \mbox{KM}}}(t)=\prod_{t_{(i)}\leq{t}}\frac{n_{i}-d_{i}}{n_{i}}=\frac{n-d_{1}}{n}\cdot\frac{n-d_{1}-d_{2}}{n-d_{1}}\cdot\cdot\cdot\cdot\cdot\frac{n-d_{1}-d_{2}-\cdot\cdot\cdot-d_{m}}{n-d_{1}-d_{2}-\cdot\cdot\cdot-d_{m-1}}=\frac{n-d_{1}-d_{2}-\cdot\cdot\cdot-d_{m}}{n}\newline
    =\frac{\#individuals\;with\;survival\;times\geq{t}}{\#individuals\;in\;the\;data\;set}=\hat S_{\mbox{e}}(t)$

2. In the absence of censoring, show that the Greenwood Formula (page 30 on note 2) can be reduced to $$\frac{\hat S_{{\tiny \mbox{KM}}}(t) \times \{ 1 - \hat S_{{\tiny \mbox{KM}}}(t)\}}{n}.$$
You might assume there are no ties among the observations.

    $Var\{\widehat{S}_{KM}(t)\}\approx{\widehat{S}_{KM}^2}(t)\cdot\sum_{t_{(i)}\leq{t}}\frac{d_i}{n_i\cdot(n_i-d_i)}\newline$
    $=(\frac{n-d_{1}-d_{2}-\cdot\cdot\cdot-d_{m}}{n})^2\cdot\sum_{t_{(i)}\leq{t}}\frac{d_i}{n_i\cdot(n_i-d_i)}\quad assume\ no\ censoring,\ follow\ from\  Q1\newline$
    $=(\frac{n-t_{(m)}}{n})^2\cdot\sum_{t_{(i)}\leq{t}}\frac{1}{n_i\cdot(n_i-1)}\quad assume\ no\ ties \newline$
    $=(\frac{n-t_{(m)}}{n})^2\cdot(\frac{1}{n(n-1)}+\frac{1}{(n-1)(n-2)}+\cdot\cdot\cdot+\frac{1}{(n-(t_{(m)}-1))(n-t_{(m)})})\quad assume\ no\ censoring\newline$
    $=(\frac{n-t_{(m)}}{n})^2\cdot((-\frac{1}{n}+\frac{1}{n-1})+(-\frac{1}{n-1}+\frac{1}{n-2})+\cdot\cdot\cdot+(-\frac{1}{n-(t_{(m)}-1)}+\frac{1}{n-t_{(m)}}))\newline$
    $=(\frac{n-t_{(m)}}{n})^2\cdot(\frac{1}{n-t_{(m)}}-\frac{1}{n})\newline$
    $=\frac{\frac{n-t_{(m)}}{n}-(\frac{n-t_{(m)}}{n})^2}{n}\newline$
    $=\frac{\widehat{S}_{KM}(t)\times\{1-\widehat{S}_{KM}(t)\}}{n}$

3. Consider the Leukemia data from the `survival` package: \label{aml}
    ```{r aml}
    library(survival)
    head(aml)
    ```
    In here, each row represent one patient. `aml` is the observed survival time, `status` is the censoring indicator (1 = event, 0 = censored), and `x` is the treatment indicator. We will ignore the treatment indicator for now.
    a. Plot the Kaplan-Meier survival curve for the data. \label{aml:surv}

    ```{r q3a}
    km <- survfit(Surv(time, status) ~ 1, data = aml)
    plot(km,main="Kaplan-Meier survival curve")
    legend(140, 1, legend=c("Kaplan-Meier","KM 95% CI"),col=c("black","black"),
    lty=c(1,2), cex=0.5)
    ```

    b. Add the Nelson-Aalen survival curve to the Kaplan-Meier plot from \eqref{aml:surv}. 

    ```{r q3b}
    cox <- coxph(Surv(time, status) ~ 1, data = aml)
    H0 <- basehaz(cox)
    plot(km,main="Kaplan-Meier and Nelson-Aalen survival curve")
    lines(H0$time, exp(-H0$hazard), 's', col = 2)
    legend(140, 1, legend=c("Kaplan-Meier","KM 95% CI", "Nelson-Aalen"),
    col=c("black","black", "red"),lty=c(1,2,1), cex=0.5)
    ```

4. The expected survival time for the Leukemia data in \#\eqref{aml} does not exist because the last observation is a censored event.
    An alternative is to lookInstead of looking at the expected survival time, an alternative is to look at the restricted mean survival time.
    Compute $E(T|T<161)$ based on the survival curve in \eqref{aml:surv}.

    $E(T|T<a)=\int_{0}^{a} t\frac{f_{T}(t)}{1-S_{T}(a)}dt+\int_{a}^{\infty} tf_{T|T<a}(t)dt=\frac{1}{1-S_{T}(a)}\int_{0}^{a} tf_{T}(t)dt+0=\frac{1}{1-S_{T}(a)}\int_0^a S_{T}(t)dt$

    ```{r q4}
    matrix(c(km$time,(km$surv-tail(km$surv, n=1))/(1-tail(km$surv, n=1))),nrow=length(km$time),
    dimnames = list(NULL, c("death_time","km_surv"))) %>%
    as.data.frame() %>%
    add_row(death_time = 0, km_surv = 1,.before = 1) %>%
    arrange(desc(km_surv)) %>%
    mutate(time_diff=lead(death_time,default = 0)-death_time) %>%
    filter(death_time<161) %>%
    summarise(expected_lifetime=sum(time_diff*km_surv))
    ```
    The conditional expected lifetime is 25.11061.

5. Let $N_i(t)$ be the number of events over time interval $(0, t]$ for the $i$th patient in \#\eqref{aml}. Let $N(t)=\sum_{i=1}^nN_i(t)$ be the aggregated counting process.
    a. Plot $N(t)$.

    ```{r q5a}
    calculate_Nt=
    aml %>% arrange(time) %>% select(-x) %>% group_by(time) %>% 
    summarize(di = sum(status)) %>% 
    mutate(Nt = cumsum(di))
    
    qplot(time, Nt, data = calculate_Nt, geom = "step",
    main="N(t) plot") +
    ylab("N(t)") + geom_point()
    ```
    b. Plot $M(t)$, where $M(t) = N(t) - \hat H(t)$ and $\hat H(t)$ is the Nelson-Aalen estimator for the cumulative hazard function. \label{Mt}
    
    * Note on \ref{Mt}: After giving some thought, I think it is more meaningful to plot $dM(t)=dN(t)- \hat h(t)dt$. Both plots will receive full credit for \ref{Mt}.

    ```{r q5b}
    na_surv_by_formula=
    aml %>% arrange(time) %>% select(-x) %>% group_by(time) %>% 
    summarize(di = sum(status), ni = length(status)) %>% 
    mutate(ni = rev(cumsum(rev(ni))), hi = di/ni,
    Nt = cumsum(di), Ht = cumsum(hi), Mt = Nt-Ht,
    dNt = di, htdt = hi, dMt = di-hi)

    qplot(time, Mt, data = na_surv_by_formula, geom = "step",
    main="M(t) plot from Nelson-Aalen estimator using the formula") +
    ylab("M(t)") + geom_point()
    
    qplot(time, dMt, data = na_surv_by_formula,
    main="dM(t) plot from Nelson-Aalen estimator using the formula") +
    geom_abline(intercept = 0, slope = 0) +
    geom_point(aes(na_surv_by_formula$time, rep(0,18)), pch=1) + ylab("dM(t)")
    
    na_surv_by_cosph=
    aml %>% arrange(time) %>% select(-x) %>% group_by(time) %>% 
    summarize(di = sum(status), ni = length(status)) %>% 
    mutate(ni = rev(cumsum(rev(ni))), hi = di/ni,
    Nt = cumsum(di), Ht = H0$hazard, Mt = Nt-Ht,
    dNt = di, htdt = diff(c(0,Ht)), dMt = di-htdt)

    qplot(time, Mt, data = na_surv_by_cosph, geom = "step",
    main="M(t) plot from Nelson-Aalen estimator using coxph") +
    ylab("M(t)") + geom_point()
    qplot(time, dMt, data = na_surv_by_cosph,
    main="dM(t) plot from Nelson-Aalen estimator using coxph") +
    geom_abline(intercept = 0, slope = 0) +
    geom_point(aes(na_surv_by_cosph$time, rep(0,18)), pch=1) + ylab("dM(t)")
    ```

    Since the estimators from the formula and the coxph are very close, they plots are almost identical.