---
title: Exam 1
author: Xinyu Zhang
output: pdf_document
---

```{R setup, include = FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```
**Instructions**

  * Due date: **Tuesday, November 13**.
  * This is a open resource exam, but you are not allowed to ask post exam questions online.
  * You are not allowed to collaborate with classmates and/or people outside of class.
  * Please circle or highlight your final answer. 
  * The toal possible point is 80.
  
Violation of this agreement will result in an **F** on this exam and it will be averaged in as a 0\%.
  
  
---

```{r load package and dataset, echo=FALSE, message=FALSE}
library(survMisc)
library(tidyverse)

data(whas100, package = "survMisc")
```

1. Use the complete *WHAS100* dataset and gender as the group indicator, 
compute the log-rank statistic, $Q$, presented as Equation (9) on page 60 of note 2 with $\omega_i = 1$.
Use $Q$ to compute a $p$-value to test the null hypothesis of $H_o: S_0(t) = S_1(t)$.
Do this **without a software package** (6 pts), and verify the calculation with `survdiff` (4 pts).

```{r gp1 and gp0 side by side}
km <- survfit(Surv(lenfol, fstat) ~ 1, data = whas100)
km.gp1=with(whas100 %>% filter(gender==1),survfit(Surv(lenfol, fstat) ~ 1))
km.gp0=with(whas100 %>% filter(gender==0),survfit(Surv(lenfol, fstat) ~ 1))

km.df=data.frame(time=km$time,n.risk=km$n.risk)
km.df.gp1=data.frame(time=km.gp1$time,d1.event=km.gp1$n.event,c1.censor=km.gp1$n.censor)
km.df.gp0=data.frame(time=km.gp0$time,d0.event=km.gp0$n.event,c0.censor=km.gp0$n.censor)

km.gender=left_join(km.df,km.df.gp1,by="time") %>%
  mutate(d1.event=replace_na(d1.event,0),c1.censor=replace_na(c1.censor,0)) %>%
  mutate(n1.risk=35-head(c(0,cumsum(d1.event+c1.censor)),-1),s1=cumprod(1-d1.event/n1.risk)) %>%
  left_join(km.df.gp0,by="time") %>%
  mutate(d0.event=replace_na(d0.event,0),c0.censor=replace_na(c0.censor,0)) %>%
  mutate(n0.risk=65-head(c(0,cumsum(d0.event+c0.censor)),-1),s0=cumprod(1-d0.event/n0.risk)) %>%
  mutate(s1=replace_na(s1,tail(s1[!is.na(s1)],1)),s0=replace_na(s0,tail(s0[!is.na(s0)],1)))

head(km.gender)
```

We put the Kaplan-Meier survival function of group 1(gender==1) and group 0(gender==0) in the same table.

```{r calculate q without package}
km.gender %>%
  mutate(e.d=n1.risk*(d1.event+d0.event)/n.risk,
    var.d=(n1.risk*n0.risk*(d1.event+d0.event)*(n.risk-d1.event-d0.event))/n.risk^2/(n.risk-1)) %>%
  summarise(q.numerator=(sum(d1.event-e.d))^2, q.denominator=sum(var.d[!is.na(var.d)])) %>%
  mutate(q=q.numerator/q.denominator)
```

```{r calculate q with survdiff}
survdiff(Surv(lenfol, fstat) ~ gender, data = whas100)
```

We see that the q calculated by both methods are the same(3.9714).

2. There are many ways to form a basis for survival curve comparison. \label{stats}
Here are some:
    a. The numerator term in $Q$ without the square:
    $$D_1 = \sum_{i = 1}^D\{d_{1i} - E(d_{1i})\}.$$
    b. The largest distance between the two curves:
    $$D_2 = \max|S_1(t) - S_0(t)|.$$
    d. The difference between the median survival times:
    $$D_3 = S_1^{-1}(0.5) - S_0^{-1}(0.5)$$
    d. The difference between the mean survival times:
    $$D_4 = \int_0^{t_{(n)}} \{S_1(u) - S_0(u)\}\,du,$$ where $t_{(n)}$ is the maximum observed survival time.
    
    \flushleft Compute each of the above statistic for the $WHAS100$ dataset (5 pts $\times4$).

```{r calculate d}
#d1
with(km.gender,sum(d1.event-n1.risk*(d1.event+d0.event)/n.risk))

#d2
with(km.gender,max(abs(s1-s0)))

#d3
with(km.gender,min(time[s1<=0.5])-min(time[s0<=0.5]))

#d4
with(km.gender,sum(head(c(1,s1),-1)*(km.gender$time-head(c(0,km.gender$time),-1))-
                        head(c(1,s0),-1)*(km.gender$time-head(c(0,km.gender$time),-1))))

```

D1, D2, D3 and D4 are 6.62, 0.3854, -818 and -432.2089 respectively.

3. The statistics computed in \eqref{stats} do not provide meaningful interpretations when standing along. 
We will use a permutation approach to test for the null hypothesis of $H_o: S_0(t) = S_1(t)$ based on these statistics.
The idea of a permutation test is simple. The general procedure can be summarized into the following steps:
    i. Compute the desired statistic based on the observed data; we will call this the observed statistic.
    ii. Permute the data under the null.
    iii. Compute the statistics for each possible permutation in Step ii.; we will call these permutation statistics.
    iv. Draw conclusion based on where the observed statistic stands among the permutation statistics.
\newline

    \flushleft
The statistics we computed in \eqref{stats} are the observed statistics in Step i.
If the null hypothesis of $H_o: S_0(t) = S_1(t)$ is true, 
then one can randomly shuffle the group indicator to generate different permutations (Step ii) and the statistics for these permutations should be similar (Step iii).
\newline

    \flushleft
    a. (5 pts$\times4$) Generate 5000 permutation and, for each of the permutation, compute the four statistics presented in \eqref{stats}.
    We will call the permutated statistics $D_{1i}^*$, $D^*_{2i}$, $D^*_{3i}$ and $D^*_{4i}$ for $i = 1, \ldots, 5000$.
    Create a histgram for these permutated statistics and print the `summary`.
    \flushleft
    b. (5 pts$\times4$) Compute the $p$-value based on these statistics by 
    $$p = 2\cdot\frac{\min(N_1, N_2)}{5000}, $$
    where $N_1 = \#\{D \ge D^*\}$, $N_2 = \#\{D\le D^*\}$, and $\#$ means the "number of", e.g., $N_1$ is the number of these permutated statistics less than or equal to the observed statistic.

```{r define d.fun and permu test,warning=FALSE}
d.fun = function(data) {
  km <- survfit(Surv(lenfol, fstat) ~ 1, data = data)
  km.gp1=with(data %>% filter(gender==1),survfit(Surv(lenfol, fstat) ~ 1))
  km.gp0=with(data %>% filter(gender==0),survfit(Surv(lenfol, fstat) ~ 1))
  
  km.df=data.frame(time=km$time,n.risk=km$n.risk)
  km.df.gp1=data.frame(time=km.gp1$time,d1.event=km.gp1$n.event,c1.censor=km.gp1$n.censor)
  km.df.gp0=data.frame(time=km.gp0$time,d0.event=km.gp0$n.event,c0.censor=km.gp0$n.censor)
  
  km.gender=left_join(km.df,km.df.gp1,by="time") %>%
    mutate(d1.event=replace_na(d1.event,0),c1.censor=replace_na(c1.censor,0)) %>%
    mutate(n1.risk=35-head(c(0,cumsum(d1.event+c1.censor)),-1),s1=cumprod(1-d1.event/n1.risk)) %>%
    left_join(km.df.gp0,by="time") %>%
    mutate(d0.event=replace_na(d0.event,0),c0.censor=replace_na(c0.censor,0)) %>%
    mutate(n0.risk=65-head(c(0,cumsum(d0.event+c0.censor)),-1),s0=cumprod(1-d0.event/n0.risk)) %>%
    mutate(s1=replace_na(s1,tail(s1[!is.na(s1)],1)),s0=replace_na(s0,tail(s0[!is.na(s0)],1)))
  
  
  d1=with(km.gender,sum(d1.event-n1.risk*(d1.event+d0.event)/n.risk))
  d2=with(km.gender,max(abs(s1-s0)))
  d3=with(km.gender,min(time[s1<=0.5])-min(time[s0<=0.5]))
  d4=with(km.gender,sum(head(c(1,s1),-1)*(km.gender$time-head(c(0,km.gender$time),-1))-
                          head(c(1,s0),-1)*(km.gender$time-head(c(0,km.gender$time),-1))))
  
  return(list(d1=d1,d2=d2,d3=d3,d4=d4))
}

d.obs=d.fun(data=whas100)

set.seed(123)
dist.d.list=
 with(whas100,
  replicate(5000,
   d.fun(data=data.frame(lenfol=lenfol,fstat=fstat,gender=sample(gender,length(gender),FALSE)))))

dist.d.matrix=matrix(unlist(dist.d.list),ncol=4,byrow=TRUE)
```

We first define funtion d.fun which calculates D1, D2, D3 and D4 from the data. Then we sample gender without replacement and create a new dataset from it. Finally, we use d.fun to calculates D1, D2, D3 and D4. The last two steps are repeated 5000 times. dist.d.matrix(5000 by 4 matrix) holds the sample distribution for D1, D2, D3 and D4.

```{r d1}
dist.d1=dist.d.matrix[,1]

hist(dist.d1)
abline(v=d.obs$d1,col="blue",lwd = 2)
summary(dist.d1)

2*min(sum(dist.d1>=d.obs$d1),sum(dist.d1<=d.obs$d1))/5000
```

P-value for D1 is 0.046.

```{r d2}
dist.d2=dist.d.matrix[,2]

hist(dist.d2)
abline(v=d.obs$d2,col="blue",lwd = 2)
summary(dist.d2)

2*min(sum(dist.d2>=d.obs$d2),sum(dist.d2<=d.obs$d2))/5000 
```

P-value for D2 is 0.536. Note that the largest survival time is a censored observation, so the permutation test doesn't center at 0.

```{r d3}
dist.d3=dist.d.matrix[,3][is.finite(dist.d.matrix[,3])]

hist(dist.d3)
abline(v=d.obs$d3,col="blue",lwd = 2)
summary(dist.d3)

2*min(sum(dist.d3>=d.obs$d3),sum(dist.d3<=d.obs$d3))/length(dist.d3) 
```

P-value for D3 is 0.1246147. Note that some sameples do not have median survival function, so we remove those D3 and adjust the denominator when calculate the p-value.

```{r d4}
dist.d4=dist.d.matrix[,4]

hist(dist.d4)
abline(v=d.obs$d4,col="blue",lwd = 2)
summary(dist.d4)

2*min(sum(dist.d4>=d.obs$d4),sum(dist.d4<=d.obs$d4))/5000 
```

P-value for D4 is 0.0496.

4. Another method to compare two survival curves is to consider a sign test. 
Suppose we have two groups of uncensored survival times:

      \flushleft Males: $x_1, x_2, \ldots, x_{n_0}$.
      \flushleft Females: $y_1, y_2, \ldots, y_{n_1}$.

      \flushleft
The sign test looks at the statistic
$$ U = \sum_{i = 1}^{n_0}\sum_{j = 1}^{n_1}\mbox{sgn}(x_i - y_j),$$
where $\mbox{sgn}(\cdot)$ is the sign function. 
In the pretense of right censoring, survival times can not be compared directly and a modified version of $U = \sum_{i = 1}^{n_0}\sum_{j = 1}^{n_1}U_{ij}$ is considered, where
$$U_{ij}=\left\{\begin{matrix}
1 & \mbox{if } x_i > y_j, y_j \mbox{ is uncersored.}\\
-1 & \mbox{if } x_i < y_j, x_i \mbox{ is uncersored.}\\ 
0 & \mbox{otherwise.} 
\end{matrix}\right.$$

    a. (5 pts) Compute $U$ for the WHAS100 dataset.
    b. (5 pts) Create a histgram for these permutated statistics and print the `summary`. Obtain a permutation $p$-value based on 5000 permutations.

```{r calculate u}
grid.1u0u=with(whas100 %>% filter(fstat==1),
               expand.grid(lenfol[gender==1],lenfol[gender==0]))
grid.1c0u=with(whas100 %>% filter((fstat==0&gender==1)|(fstat==1&gender==0)),
               expand.grid(lenfol[gender==1],lenfol[gender==0]))
grid.1u0c=with(whas100 %>% filter((fstat==1&gender==0)|(fstat==0&gender==1)),
               expand.grid(lenfol[gender==1],lenfol[gender==0]))

##u
with(grid.1u0u,sum(Var1>Var2)-sum(Var1<Var2))+
with(grid.1c0u,sum(Var1>Var2))+
with(grid.1u0c,-sum(Var1<Var2))
```

U is 256.

```{r define u function and permu test}
u.fun = function(data) {
  grid.1u0u=with(data %>% filter(fstat==1),
                 expand.grid(lenfol[gender==1],lenfol[gender==0]))
  grid.1c0u=with(data %>% filter((fstat==0&gender==1)|(fstat==1&gender==0)),
                 expand.grid(lenfol[gender==1],lenfol[gender==0]))
  grid.1u0c=with(data %>% filter((fstat==1&gender==0)|(fstat==0&gender==1)),
                 expand.grid(lenfol[gender==1],lenfol[gender==0]))
  
  u=with(grid.1u0u,sum(Var1>Var2)-sum(Var1<Var2))+
    with(grid.1c0u,sum(Var1>Var2))+
    with(grid.1u0c,-sum(Var1<Var2))
  
  return(u)
}

u.obs=u.fun(whas100)
```

```{r u hist}
set.seed(123)
dist.u=
 with(whas100,
  replicate(5000,
   u.fun(data=data.frame(lenfol=lenfol,fstat=fstat,gender=sample(gender,length(gender),FALSE)))))

hist(dist.u)
abline(v=u.obs,col="blue",lwd = 2)
summary(dist.u)

2*min(sum(dist.u>=u.obs),sum(dist.u<=u.obs))/5000 
```

P-value is 0.2052.
